### 2017-11-14 submission/manuscript.Rmd
add updated numbers

#Read study data files
stool_n <- read_csv("../data/process/tables/stool_study_n_analyzed.csv")
tissue_n <- read_csv("../data/process/tables/tissue_study_n_analyzed.csv")

@@

**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer (CRC). Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r sum(filter(tissue_n, study == "flemer")[, c("control", "adenoma", "cancer")]) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.

@@

Colorectal cancer (CRC) is a growing world wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation in mouse models. There has also been numerous case/control studies investigating the microbiota in both adenoma and carcinoma. Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. Many of the studies along with the current meta-analysis focus on identifying biomarkers or individual microbes but do not critically investigate the community role in the disease.

Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] we expand both the breadth and scope of the previous meta-analysis; to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes across disease (control to adenoma to carcinoma) and if it results in an increased relative risk for adenoma or carcinoma. Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models we analyzed whether the full community or only the CRC associated genera resulted in better model classification area under the curve (AUC). Finally, we examined at what effect size and sample n the studies that were used were adequately powered for.

>>>>

#Read study data files
stool_n <- read_csv("../data/process/tables/stool_study_n_analyzed.csv")
tissue_n <- read_csv("../data/process/tables/tissue_study_n_analyzed.csv")
flemer_data <- read_tsv("../data/process/flemer/flemer.metadata")

@@

**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer (CRC). Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.

@@

Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. Many of the studies, along with the previous meta-analysis, focus on identifying biomarkers or individual microbes but do not critically investigate the role the community has in CRC.

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC). Since the changes in community were subtle for adenoma we also examined what effect size and sample size the studies that were used were adequately powered for.

### 2017-11-14 submission/manuscript.tex
add updated numbers

\textbf{Background.} An increasing body of literature suggests that
there is a role for the microbiota in colorectal cancer (CRC). Important
drivers within this context have ranged from individual microbes to the
whole community. Our study expands on a recent meta-analysis
investigating microbial biomarkers for CRC by testing the hypothesis
that the bacterial community is an important driver of both early
(adenoma) and late (carcinoma) stage of disease. To test this hypothesis
we examined both feces (n = 1737) and tissue (492 total samples from 400
individuals) across 14 different studies.

@@

Colorectal cancer (CRC) is a growing world wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in polyp formation in mouse models. There has also been
numerous case/control studies investigating the microbiota in both
adenoma and carcinoma. Recently, a meta-analysis was published
investigating whether specific biomarkers could be consistently
identified using multiple data sets {[}11{]}. Many of the studies along
with the current meta-analysis focus on identifying biomarkers or
individual microbes but do not critically investigate the community role
in the disease.

Using both fecal (total individuals = 1737) and tissue samples (total
samples = 492) totalling over 2229 total samples across 14 studies
{[}12--25{]} we expand both the breadth and scope of the previous
meta-analysis; to investigate whether the bacterial community is an
important risk factor for both adenoma and carcinoma. To accomplish this
we first assessed whether the diversity changes across disease (control
to adenoma to carcinoma) and if it results in an increased relative risk
for adenoma or carcinoma. Next, we assessed how typical CRC associated
genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
risk of adenoma or carcinoma. Third, using Random Forest models we
analyzed whether the full community or only the CRC associated genera
resulted in better model classification area under the curve (AUC).
Finally, we examined at what effect size and sample n the studies that
were used were adequately powered for.

>>>>

\textbf{Background.} An increasing body of literature suggests that
there is a role for the microbiota in colorectal cancer (CRC). Important
drivers within this context have ranged from individual microbes to the
whole community. Our study expands on a recent meta-analysis
investigating microbial biomarkers for CRC by testing the hypothesis
that the bacterial community is an important driver of both early
(adenoma) and late (carcinoma) stage of disease. To test this hypothesis
we examined both feces (n = 1737) and tissue (492 total samples from 350
individuals) across 14 different studies.

@@

Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
in which the microbiota has been purported to play an active role in
disease pathogenesis {[}2{]}. Numerous studies have shown the importance
of both individual microbes {[}3--7{]} and the overall community
{[}8--10{]} in polyp formation using mouse models of CRC. There have
also been numerous case/control studies investigating the microbiota in
the formation of both adenoma and carcinoma. Recently, a meta-analysis
was published investigating whether specific biomarkers could be
consistently identified using multiple data sets {[}11{]}. Many of the
studies, along with the previous meta-analysis, focus on identifying
biomarkers or individual microbes but do not critically investigate the
role the community has in CRC.

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12--25{]}, we expand both the breadth and scope of the previous
meta-analysis to investigate whether the bacterial community is an
important risk factor for both adenoma and carcinoma. To accomplish this
we first assessed whether the diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased
relative risk (RR) for adenoma or carcinoma. Next, we assessed how
typical CRC associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
risk of adenoma or carcinoma. Third, using Random Forest models, we
analyzed whether the full community or only the CRC associated genera
resulted in better model classification based on the area under the
curve (AUC). Since the changes in community were subtle for adenoma we
also examined what effect size and sample size the studies that were
used were adequately powered for.
